Insight Into Lumbar Back Pain \subtitle{What the Lumbar Spine Tells About Your Life}

### 2014-09-03 draft.tex
Crushed todo list

\subsection{Image Data}
\begin{figure}[!h]
  %\vspace{-0.2cm}
  \centering
  \includegraphics[width=0.5\textwidth]{figures/centerline}
  \caption{
	(a) left: Tetrahedron-based finite element model (FEM) of the lumbar spine, capturing the L1-L5 vertebrae as well as the lumbar spine canal.
	%%
	(a) right: FEM aligned to a MRI data set.
	%%
	The purple dashed line represents the centerline, describing the lumbar spine canal with 92 points.
	%%
	(b) Using the Frenet formulas, we extract the weighted sum of curvature and torsion for all 92 points (white dashes).
	%%
	The curvature angle ($\alpha$) is calculated using the top, bottom and middle point.
	%%
	We also extracted the variables for each projection axis to assess their information gain.
	%(b) Using the Frenet formulas, we extract the curvature, torsion and the curvature angle from the lumbar spine canal center line.
	}
  \label{fig:centerline}
\end{figure}
Four trained radiology technicians acquired the medical image data for each subject in a standardized way using a 1.5~Tesla MRI scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany).
%%
The spine protocol consisted of a sagittal T1-weighted turbo-spin-echo sequence ($1.1\times1.1\times4.0~mm$ voxels) and a sagittal T2-weighted turbo-spin-echo sequence ($1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.
%%

@@

We have to assess the model accuracy to extract key figures from it.
%%
The detection model depicts the vertebrae positions, but lacks detailed information about their volume.
%%
It captures reliable information about spine canal curvature.
%%
In a previous work (\cite{Klemm2013VMV}), we extracted a centerline representation of the lumbar spine canal from the detection model (Fig.~\ref{fig:centerline}~a).
%%
Using the Frenet formulas, we calculated the following metrics from the model (Fig.~\ref{fig:centerline}~b) \cite{Frenet}:
%%
%Thus, we extracted the following metrics from the model:
%%
%\paragraph{Curvature}
\begin{itemize}
	% \item The curvature of the centerline is calculated using the summed mean curvature between adjacent points:
% %%
% $\sum \frac{curvature_i}{curvature_all}$.
	\item \emph{Curvature} is calculated as weighted sum of curvature between all adjacent points describing the centerline: $\sum \frac{curvature_i}{curvature_{all}}$.
%%
	\item \emph{Torsion} (deviation of a curve from its current course) is calculated as weighted sum of torsion between all adjacent points describing the centerline: $\sum \frac{torsion_i}{torsion_{all}}$.
	\item \emph{Curvature angle $\alpha$} is defined by the middle point of the spine canal centerline as \emph{vertex} and the line between middle point and top/bottom point as \emph{sides}.
%%
\end{itemize}
%%
These figures are also extracted in the sagittal, coronal and transversal projection of the model.
%%
We assess the information gain of each dimension using these projections.
%%
This gives us a total of 9 image-derived parameters.

@@

  \includegraphics[width=0.5\textwidth]{figures/age-gender}

@@

Another observation during the analysis is very large decision trees achieving low error scores.
%%
Their complexity impose overfitting to the data and they are not suitable for extracting universal rules.
%%
\subsection{Proposed Features}
\label{sec:ProposedFeatures}
\begin{figure}[!h]
  %\vspace{-0.2cm}
  \centering
  \includegraphics[width=0.5\textwidth]{figures/proposed-features}
  \caption{
	Proposed features.
	%%
	(a) dented vertebraes are a sign of heavy stress and bone absorption.
	%%
	(b, c) the spine canal shape can be used to characterize scoliosis (curvature sagittal) and lordosis (curvature coronal) as well as the \emph{Cobb-Angles} \cite{Cobb}.
	%%
	(c) Spine canal thickness is associated with herniated disks.
	%%
	}
  \label{fig:proposedFeatures}
\end{figure}
%%
\noindent Pathological deformation of the lumbar spine is usually the last stage and is associated with very strong lower back pain \com{Zitat Frau Hegenscheid fragen}.
%%
We have to capture earlier signs of pathological change in order to derive better predictive models.
%%

@@

As described in a prior work (\cite{Klemm2014VIS}), the suggestion of potentially interesting features when analyzing a condition is an important aspect of visual analytics in epidemiology.
%%
We can achieve this also using the presented method by trying to describe several (dichotomous) target values with a decision tree constructed from all available data.
%%
This allows both to assess the discriminative power of the data set regarding the target variables, as well as the most important parameters for the classification.
%Except for gender in the age-plot, no variable could be explained with reasonable accuracy for age or body size.

>>>>

\subsection{Image Data}
\begin{figure}[!h]
  %\vspace{-0.2cm}
  \centering
  \includegraphics[width=0.475\textwidth]{figures/centerline}
  \caption{
	(a) left: Tetrahedron-based finite element model (FEM) of the lumbar spine, capturing the L1-L5 vertebrae as well as the lumbar spine canal.
	%%
	(a) right: FEM aligned to a MRI data set.
	%%
	The purple dashed line represents the centerline, describing the lumbar spine canal with 92 points.
	%%
	(b) Using the Frenet frame, we extract the weighted sum of curvature and torsion for all 92 points (white dashes).
	%%
	The curvature angle ($\alpha$) is calculated using the top, bottom and middle point.
	%%
	We also extracted the variables for each projection axis to assess their information gain.
	%(b) Using the Frenet formulas, we extract the curvature, torsion and the curvature angle from the lumbar spine canal center line.
	}
  \label{fig:centerline}
\end{figure}
Four trained radiology technicians acquired the medical image data for each subject in a standardized way using a 1.5~Tesla MRI scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany).
%%
The spine protocol consisted of a sagittal T1-weighted turbo-spin-echo sequence ($1.1\times1.1\times4.0~mm$ voxels) and a sagittal T2-weighted turbo-spin-echo sequence ($1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.
%%

@@

We have to assess the model accuracy to extract key figures from it.
%%
The detection model depicts the vertebrae positions, but lacks detailed information about their volume.
%%
It captures reliable information about spine canal curvature.
%% BLIND
%In a previous work (\cite{Klemm2013VMV}), we extracted a centerline representation of the lumbar spine canal from the detection model (Fig.~\ref{fig:centerline}~a).
The centerline representation of the lumbar spine canal was extracted from the detection model (Fig.~\ref{fig:centerline}~a) using the approach of \cite{Klemm2013VMV}.
%%
Using the Frenet frame, we calculated the following metrics from the model (Fig.~\ref{fig:centerline}~b) \cite{Frenet}:
%%
%Thus, we extracted the following metrics from the model:
%%
%\paragraph{Curvature}
\begin{itemize}
	% \item The curvature of the centerline is calculated using the summed mean curvature between adjacent points:
% %%
% $\sum \frac{curvature_i}{curvature_all}$.
	\item \emph{Normalized Curvature} is calculated as weighted sum of curvature between all adjacent points describing the centerline: $\sum_{i=1}^I \frac{curvature_i}{i}$. We refer to total curvature from now on as \emph{curvature}.
%%
	\item \emph{Normalized Torsion} (deviation of a curve from its current course) is calculated as weighted sum of torsion between all adjacent points describing the centerline: $\sum_{i=1}^I \frac{torsion_i}{i}$. We refer to total torsion from now on as \emph{torsion}
	\item \emph{Curvature angle $\alpha$} is defined by the middle point of the spine canal centerline as \emph{vertex} and the line between middle point and top/bottom point as \emph{sides}.
%%
\end{itemize}
%%
These figures are also extracted in the sagittal, coronal and transversal projection of the model.
%%
We assess the information gain of each dimension using these projections.
%%
This gives us a total of 9 image-derived parameters.

@@

  \includegraphics[width=0.475\textwidth]{figures/age-gender}

@@

Another observation during the analysis is very large decision trees achieving low error scores.
%%
Their complexity impose overfitting to the data and they are not suitable for extracting universal rules.
%%
\subsection{Proposed Features}
\label{sec:ProposedFeatures}
\begin{figure}[!h]
  %\vspace{-0.2cm}
  \centering
  \includegraphics[width=0.475\textwidth]{figures/proposed-features}
  \caption{
	Proposed features.
	%%
	(a) dented vertebraes are a sign of heavy stress and bone absorption.
	%%
	(b, c) the spine canal shape can be used to characterize scoliosis (curvature sagittal) and lordosis (curvature coronal) as well as the \emph{Cobb-Angles} \cite{Cobb}.
	%%
	(c) Spine canal thickness is associated with herniated disks.
	%%
	}
  \label{fig:proposedFeatures}
\end{figure}
%%
\noindent Pathological deformation of the lumbar spine is usually the last stage and is associated with very strong lower back pain \com{Zitat Frau Hegenscheid fragen}.
%%
We have to capture earlier signs of pathological change in order to derive better predictive models.
%%

@@

% BLIND
%As described in a prior work (\cite{Klemm2014VIS}), the suggestion of potentially interesting features when analyzing a condition is an important aspect of visual analytics in epidemiology.
Suggesting potentially interesting features when analyzing a condition is an important aspect of visual analytics in epidemiology \cite{Klemm2014VIS}.
%%
We can achieve this also using the presented method by trying to describe several (dichotomous) target values with a decision tree constructed from all available data.
%%
This allows both to assess the discriminative power of the data set regarding the target variables, as well as the most important parameters for the classification.
%Except for gender in the age-plot, no variable could be explained with reasonable accuracy for age or body size.
